---
title: "Lab 07 - Random forests and bagging"
author: '[your name here]'
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

Brief description
=================

In these two weeks we covered tree-based classification methods,
including ensemble classifiers such as bagging and random forests. You
also learned how to apply boosting on classification trees.

In this lab, we will perform random forest on a heart disease data set,
in which 13 attributes (features) including age, sex, cholesterol/blood
sugar level etc. are used to predict the heart disease status (column
14). The training and test data sets are available on *Connect*, and a
description can be found at
<http://archive.ics.uci.edu/ml/datasets/Heart+Disease>.

```{r setup, warning=FALSE, message=FALSE}
library(randomForest)
tr <- readRDS("heart_train.RDS")
te <- readRDS("heart_test.RDS")
```

Questions
=========

A classification tree can be sensitive to the input data. Bagging or
*bootstrap aggregation* tries to produce more stable predictions by
repeatedly fitting trees to bootstrapped samples (i.e., sampling from
the original data set with replacement). The final prediction is the
majority class based on the $K$ trees, where $K$ is the number of
bootstrap samples trained. By
making a single change to bagging --- that in each split of the tree we
restrict the algorithm to choosing among only $m$ randomly selected
features (smaller than $p$, the total number of features), we obtain the
random forest procedure that reduces the correlation among the trees.

We use the `randomForest()` function in the package of the same name to
fit a classification tree. Use the following lines to fit a random
forest to the training data with 1,501 trees:

The "`...`" part is for you to fill in.

**Q1** How many features are randomly selected as possible
candidates in each split (i.e., the value of $m$ in the above
description)? **Hint**: See `?randomForest()`.
    
    
::: {.solbox data-latex=""}    
```{r q1-sol}
set.seed(200)


```

:::


**Q2** Plot the fitted object. Do you think using 1,501 trees
    is sufficient? Why?\

::: {.solbox data-latex=""}

```{r q2-sol}
# some code

```


:::


**Q3** Using the `varImpPlot()` function, write down the 3 most
important variables in terms of their contribution to the decrease
in homogeneity.

::: {.solbox data-latex=""}

```{r q3-sol}
# some code

```
    

:::

**Q4** What is the out-of-bag error rate for the random forest fit in **Q1**?


::: {.solbox data-latex=""}
Some text.

:::

**Q5** What is the value of $m$ (number of candidate variables
in each split) that corresponds to bagging? Using this value of $m$,
perform bagging by including also the argument `mtry=m` in
`randomForest`. What is the OOB error using bagging? **[Again, make sure you run `set.seed(200)` before running `randomForest()`.]**

::: {.solbox data-latex=""}
The value of $m$ for bagging is the number of predictors so `r ncol(tr)-1`.

```{r q5-sol}
set.seed(200)


```

:::

**Q6** Obtain the misclassification rate based on the test
set, using the random forest fit in **Q1**. Do you expect this value
to be similar to the OOB error rate? Why?


::: {.solbox data-latex=""}
```{r q6-sol}
# some code

```

:::


**Q7** In the random forest fitted in **Q1**, how many
bootstrapped samples (out of 1,501) do **not** contain the third
observation? How many of these trees predict this observation (which
is OOB to them) as "Y"? **Hint**: You can obtain this information from the fitted `randomForest` object.

::: {.solbox data-latex=""}
Some text.

:::


**Q8** Notice that the random forest OOB error is lower than
the bagging OOB error for this heart rate data set. Provide two
advantages of random forests over bagging for decision trees.
**Hint(s)**: Which of trees from either fits will be more independent
from each other: the ones from a random forest or the ones from
bagging? Which method favors the use of more features?

::: {.solbox data-latex=""}
Some text.

:::
