---
title: "Lab 08 - Boosting"
author: '[Sicily Xie]'
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: false
    toc: FALSE
---

Review
======

-   **Boosting** is another method to "train" a classifier. The concept
    is to reweigh observations after each model fitting based on the
    prediction accuracy for that observation as well as the overall
    strength of that fitting. This fit-and-reweigh procedure is iterated
    and the results from the fits are aggregated to produce the final
    estimation/classification rule.

-   The procedure for **Adaboost** (adaptive boosting) for binary
    response $Y_{i}\in\{-1,1\},\mbox{ }i=1,\ldots,n$ can be understood
    as follows:

    1.  Set $j=1$. Build a classifier $T_{j}(\boldsymbol{x})$ using
        **equal weight $w_{i}=1/n$** for all observations. In this
        course we have always been fitting models this way.

    2.  After obtaining the fit, identify two quantities:

        1.  A measure of the performance of the classifier $T_{j}$ in
            (1), i.e. $\alpha_{j}$ in the lecture slides. Large values
            of $\alpha_{j}$ indicate a strong classifier with low
            misclassification rate.

        2.  For each observation, the indicator of whether it has been
            misclassified by $T_{j}$, i.e.
            $I\left(y_{i}\ne T_{j}(\boldsymbol{x}_{i})\right)$.

        Now reweigh each observation: New
        $w_{i}=\mbox{Old }w_{i}\times\exp\left\{ \alpha_{j}I\left(y_{i}\ne T_{j}(\boldsymbol{x}_{i})\right)\right\}$.
        The idea here is that the weights of correctly classified
        observations remain unchanged, while **the weights for
        misclassified observations change according to the strength of
        the classifier $T_{j}$** --- if $T_{j}$ has low (high)
        misclassification rate, the misclassified observations will be
        adjusted to have much larger (smaller) weights.

    3.  $j$ is incremented and the above steps are run again, replacing
        the old weights by the new weights. Note how this new classifier
        will adapt to the data --- since more weight is given to
        observations missed by a previous strong classifier, **this new
        classifier will put more emphasis on these hard-to-classify
        points** in an attempt to improve the overall fit.

    4.  With the above steps being run $K$ times, the final classifier
        is given by the sign of
        $\sum_{j=1}^{K}\alpha_{j}T_{j}(\boldsymbol{x})$, which takes
        into account the relative strengths of the classifiers. Strong
        classifiers obviously have heavier weights.

    This is again a general procedure that is applicable to many
    classifiers, e.g. logistic classifier, LDA, trees etc.

-   From the form of the final classifier it is easy to see that this
    algorithm fits an **additive model** in a **forward fashion**, i.e.
    upon fitting the $k$th classifier, *every* classifier before and
    including the $k$th is used to build the final classifier. It can
    also be shown that Adaboost attempts to minimize the **exponential
    loss**, the function of which is
    $$L\left(y,G(\boldsymbol{x})\right)=\exp\{-yG(\boldsymbol{x})\}$$
    where $y\in\{-1,1\}$ is the observed response and
    $G(\boldsymbol{x})$ is the classifier.

-   The exponential loss function penalizes bad mistakes much more
    heavily than the 0-1 loss function does. Suppose $y=1$ while two
    classifiers $G_{1}$ and $G_{2}$ give $-1$ and $-5$ respectively. In
    each case the 0-1 loss assigns a value of 1 (mismatch in sign), but
    the exponential loss is $e^{1}\approx2.7$ for $G_{1}$ and
    $e^{5}\approx148$ for $G_{2}$! Also note that the exponential loss
    is non-zero for correct classifications, unlike the 0-1 loss
    function.

-   The population solution for $G(\boldsymbol{x})$ that minimizes the
    exponential loss can be shown to be
    $$\frac{1}{2}\log\left(\frac{\mathbb{P}\left(Y=1|\boldsymbol{X}=\boldsymbol{x}\right)}{\mathbb{P}\left(Y=-1|\boldsymbol{X}=\boldsymbol{x}\right)}\right)\label{eq:form}$$
    as a function of $\boldsymbol{x}$. Since we are using an additive
    model to estimate [\[eq:form\]](#eq:form){reference-type="eqref"
    reference="eq:form"}, the flexibility of the classifiers we use and
    the underlying mechanism of data generation will determine our
    ability to fit the model well.

Questions
=========

In this lab we explore the power of the boosting algorithm and also try
to find out when this method could fail.

Boosting with a "good" data set
-------------------------------

Use the file `boston` data in this part. This data set is derived from
the Boston Housing data. The class label is `crime` and the two
predictors are `lon` and `lat`. We will try to classify neighbourhoods
with high crime rate using the spatial coordinates. You will need the
`{adabag}` package for the functions below.

```{r setup, message=FALSE, warning=FALSE}
library(adabag)
library(ggplot2)
bos <- readRDS("boston.RDS")
bad_boost <- readRDS("bad_boost.RDS")
```

We set the classifiers $T_{j}$ to be **stumps** --- a stump
is a tree with only one split. To specify this, run the following line:

```{r}
stump <- rpart.control(cp = -1, maxdepth = 1, minsplit = 0, xval = 0)
```

**Q1** Using the function `boosting()`, perform boosting on this data set with
150 trees (iterations). What is the misclassification error rate? Display the confusion matrix.

**Important notes --- read before you fit!**  

1. In this part set `boo=FALSE` so that the original sample is used in all trees.
1. Make use of the `control` parameter to incorporate the `stump` settings you specified above.
1. Run the line `set.seed(123)` right before running the boosting algorithm in R.

::: {.solbox data-latex=""}
```{r q1-sol}
set.seed(123)
model <- boosting(class~lon+lat, data=bos, boo=FALSE, mfinal=150,control=stump)
pre <- predict(model,newdata = bos)
confus <- pre$confusion
err <- pre$error
print(confus)
print(err)
```
The misclassification rate is 0.1007905.


:::

**Q2** Identify and plot the misclassified observations for the
boosting algorithm up to the $i$th tree, where $i=1, 10, 25,50,100$ and
$150$. You may use the provided `plot_boost()` function to generate your plots. **In the output, include the plot for $i=150$ only.** 

**Hint**: The `predict()` function in this package requires the
vector of responses (as a factor) to be present. Which parameter of `predict`
controls the number of iterations to use?

```{r plot-boost}
#'  Plot the results of a fitted boosting object
#'
#'  This function plots the class and classification error of a fitted boosting
#'  object to the boston housing data
#'  @param boost fitted boosting object
#'  @param dat data used to fit the boosting object
#'  @param ntree the number of trees of the boosting object to be used in the prediction
plot_boost <- function(boost, dat, ntree = length(boost$tree)) {
  pr <- predict(boost, newdata = dat, newmfinal = ntree)
  dat$boost.pred <- pr$class
  dat$error <- dat$boost.pred != dat$class
  ggplot(data = dat) +
    geom_point(aes(x = lon, y = lat, col = boost.pred, shape = error), size=3.5) +
    scale_shape_manual(values = c(1, 4)) +
    theme_bw()
}
```


::: {.solbox data-latex=""}
```{r hidden-plots, eval=FALSE}
# some code
for (i in c(1, 10, 25,50,100,150)) {
pred <- predict(model,newdata = bos,newmfinal=i)
print(paste0('tree:',i,',error:',pred$error))
}
```

```{r boost-150}
# some code
plot_boost(boost=model,dat=bos,ntree=150)
```
:::

Boosting with a "bad" data set
------------------------------

**Q3** Repeat **Q1** and **Q2** using the 
`bad_boost` data. What is the misclassification error rate? Does
increasing the number of iterations help? **Be sure to set the seed again.**

::: {.solbox data-latex=""}
```{r q3-sol}
set.seed(123)
model2 <- boosting(class ~ lon + lat, data = bad_boost, boos = FALSE, mfinal = 150, control = stump)
predict(model2, bad_boost)$confusion
predict(model2, bad_boost)$error
```
The misclassification error rate is 0.22.
No, increasing the number of iterations does not help.

```{r more-hidden-plots, eval=FALSE}
# some code
for (i in c(1, 10, 25,50,100,150)) {
pred <- predict(model2,newdata = bad_boost,newmfinal=i)
print(paste0('tree:',i,',error:',pred$error))
}
```

```{r bad-boost-150}
# some code
plot_boost(boost=model2,dat=bad_boost,ntree=150)
```


:::
    

**Q4** Now allow the trees to grow a bit deeper by setting
`maxdepth=2` in the `rpart.control` command while keeping other
parameters fixed at those in **Q1**. Let the algorithm iterate for
100 times. What do you observe?

::: {.solbox data-latex=""}
```{r q4-sol}
set.seed(123)
stump.new <- rpart.control(cp = -1, maxdepth = 2, minsplit = 0, xval = 0)
model3 <- boosting(class~lon+lat,data=bad_boost,mfinal=100,boo=FALSE,control=stump.new)
predict(model3, bad_boost)$confusion
predict(model3, bad_boost)$error
```
The misclassification error rate now is 0. 
The misclassification error decreases as the number of iterations increases, and converges to zero when
the number of iterations get close to 80.

:::


What's the difference?
----------------------

The response variables in the data set `bad_boost` were in fact
generated through the following relationship:
$$\frac{1}{2}\log\left(\frac{\mathbb{P}\left(Y=a|\boldsymbol{X}=\boldsymbol{x}\right)}{1-\mathbb{P}\left(Y=a|\boldsymbol{X}=\boldsymbol{x}\right)}\right)=\left[(x_{2}-2)_{+}-(x_{1}+1)_{+}\right](1-x_{1}+x_{2})$$
where $(z)_{+}=\max(z,0)$. Recall that we attempt to estimate
$$\frac{1}{2}\log\left(\frac{\mathbb{P}\left(Y=a|\boldsymbol{X}=\boldsymbol{x}\right)}{1-\mathbb{P}\left(Y=a|\boldsymbol{X}=\boldsymbol{x}\right)}\right)$$
using a **linear combination** of the individual classifiers
$T_{j}(\boldsymbol{x})$ in boosting.

**Q5** $We observed that boosting stumps yields good results on
the first data set but not the second. Why is it so?
    
::: {.solbox data-latex=""}
Since boosting performs well when combing several weak learners into strong learners. 
If there aren’t enough interactions (second data set), boosting would not perform well.

:::

![Boston Housing Data - Crime Rate](Boston.pdf){width="1\\linewidth"}
